\textbf{Differences in COVID-19 vaccination in the province of Ontario across Health Regions and socio-economic strata}

### 2023-07-02 manuscript_v6/01-background.qmd
updated methods

Reasons given for this inequality in vaccination uptake have included medical mistrust due to systemic medical racism [@stoler2021;@mosby2021], mistrust in vaccines [@willis2021], and the existence of health misinformation and disinformation [@bogart2021;@mosby2021;@freeman2020]. However, it is important also to consider that vaccination uptake can be influenced by geographical (spatial) factors. In this regard, differences in COVID-19 vaccination rates have been associated with varied regional attitudes towards vaccination [@malik2020], lack of vaccine access in areas where underrepresented groups reside[@bogoch2022], as well as an absence of prioritization of areas inhabited by vulnerable groups [@nguyen2021]. Other studies have also shown heterogeneity in vaccine uptake within small governmental administrative units such as counties [@mollalo2021; @yang2022; @tiu2022; @bhuiyan2022], indicating that accounting for geographical differences in vaccination can help predict patterns of booster uptake [@wood2022].

However, the majority of such analyses have been carried out outside of Canada, and existing studies have been focused in certain cities (such as Toronto [@choi2021], or Montreal [@mckinnon2021]), or have explored differences at a province-wide level [@guay2022]. **Canada’s pandemic response goals, were to reduce morbidity/mortality and to reduce societal disruption from COVID-19**.

- How does this integrate with vaccination? We need to study vaccination because 
-

- -

and that consequently, can help identify disparities that need to be addressed within specific areas in each province.

This need is especially important in the case of Ontario, the most populated province of Canada. Between 2007 and 2019, Ontario managed healthcare access to its inhabitants using 14 intra-provincial divisions called the Local Health Integration Networks (LHINs), which aimed to provide an integrated health system for the province. However, this approach was complex, bureaucratic, and resulted in excessive expenditures, disparities in mortality rates, the deterioration of certain performance indicators such as wait times and hospital readmissions, fragmented electronic health systems, and inequities in health services access [@tsasis2012;@muratov2018;@dong2022;@lysyk2015;@lysyk2016]. Therefore, with the intent of better organizing and delivering care in late 2019 the provincial government eliminated the LHINs and incorporated the areas covered by them into six larger Health Regions (North East, North West, Central, Toronto, West, and East), which are managed by the new government agency Ontario Health (OH) [@dong2022].

On the other hand, public health in Ontario is administered by Public Health Ontario (PHO), a government agency established in late 2007 and currently composed of 34 Public Health Units (PHUs) that cover the entire geography of the province. Although these PHUS were commissioned with leading the distribution of the COVID-19 vaccine within the areas they covered [@rotenberg2021], there was significant overlap between PHO and OH in the actual vaccine rollout, and in many instances vaccines were administered in vaccination clinics ran by members of OH working alone or in conjunction with members of PHO [@ashcroft2023].

- add about intersection of Health Regions and PHUs. How vaccination presented a challenge as vaccine was ultimately administered in facilites managed by OHT?

https://www.mcmasterforum.org/docs/default-source/product-documents/rapid-responses/examining-the-intersections-between-ontario-health-teams-and-public-health.pdf?sfvrsn=69caa8c9_11

Because of the relatively recent adoption of the Health Region model and its alignment with the onset of the COVID-19 pandemic, there is a need to analyze if there are ongoing disparities in health access under this approach that need to be addressed before they are exploited by a new disease or public health threat. In this regard, previous research has highlighted disparities in the level of activity of each Health Region [@sethuram2023]. Therefore, analyzing differences in vaccination uptake within the Health Regions and can help identify which socio-demographic groups are the most vulnerable and what areas of the province deserve special attention by decision-makers.

Therefore, in this study we aim to understand the differences in vaccination uptake between the different Health Regions of Ontario between October of 2021 and January of 2022. By including socio-economic factors in our analysis, we aimed to identify in which groups these differences were significant in order to provide an assessment of the current state of healthcare access in Ontario.

>>>>

Reasons given for this inequality in vaccination uptake have included medical mistrust due to systemic medical racism [@stoler2021;@mosby2021], mistrust in vaccines [@willis2021], and the existence of health misinformation and disinformation [@bogart2021;@mosby2021;@freeman2020]. However, it is important also to consider that vaccination uptake can be influenced by geographical (spatial) factors. In this regard, differences in COVID-19 vaccination rates have been associated with varied regional attitudes towards vaccination [@malik2020], lack of vaccine access in areas where underrepresented groups reside[@bogoch2022], as well as an absence of prioritization of areas inhabited by vulnerable groups [@nguyen2021]. Other studies have also shown heterogeneity in vaccine uptake within small governmental administrative units such as counties [@mollalo2021; @yang2022; @tiu2022; @bhuiyan2022], indicating that accounting for geographical differences in vaccination can help predict patterns of booster uptake [@wood2022]. However, there is a limited amount of studies that have analyzed geographical differences in vaccination in Canada. Existing studies in this area have focused on differences within certain cities, such as Toronto [@choi2021], or Montréal [@mckinnon2021]), or have explored differences between provinces [@guay2022], but to our knowledge, there are no studies that analyze  differences in vaccination uptake at the provincial level. Such need is specially important in the context of Canada's pandemic response goals, which have been to minimize serious illness and deaths while minimizing societal disruption [@naci2023]. Analyzing differences in vaccination uptake within the provinces can aid to identify inequalities that might exist and that need to be addressed before the advent of another pandemic.

<!-- However, in the context of Canada's pandemic response goals, which have been to minimize serious illness and deaths while minimizing societal disruption [@naci2023], there is an ongoing need to analyze the existence of geographical disparities in vaccination uptake within the Canadian provinces and to identify the socio-demographic groups affected, as this can help identify inequalities that might exist and that need to be addressed before the advent of a future pandemic. -->

<!-- **Canada’s pandemic response goals, were to reduce morbidity/mortality and to reduce societal disruption from COVID-19**.   -->

<!-- - How does this integrate with vaccination? We need to study vaccination because  -->
<!-- -  -->

<!-- - - -->

<!-- and that consequently, can help identify disparities that need to be addressed within specific areas in each province. -->

<!-- However, in the context of Canada's pandemic response goals, which have been to minimize serious illness and deaths while minimizing societal disruption [@naci2023], there is an ongoing need to analyze the existence of geographical disparities in vaccination uptake within the Canadian provinces and to identify the socio-demographic groups affected, as this can help identify inequalities that might exist and that need to be addressed before the advent of a future pandemic.  -->

This need is especially important in the case of Ontario, the most populated province of Canada, and which as has a complex healthcare system. Between 2007 and 2019, Ontario managed healthcare access to its inhabitants using 14 intra-provincial divisions called the Local Health Integration Networks (LHINs), which aimed to provide an integrated health system. However, this approach was complex, bureaucratic, resulting in excessive expenditures, disparities in mortality rates, the deterioration of certain performance indicators (such as wait times and hospital readmissions), fragmented electronic health systems, and inequities in health services access [@tsasis2012;@muratov2018;@dong2022;@lysyk2015;@lysyk2016]. With the intent of better organizing and delivering care, in late 2019 the provincial government eliminated the LHINs and incorporated the areas covered by them into six larger Health Regions (North East, North West, Central, Toronto, West, and East), which are managed by a new government agency, Ontario Health (OH) [@dong2022].

On the other hand, public health in Ontario is administered by Public Health Ontario (PHO), a government agency established in late 2007 and that is currently composed of 34 Public Health Units (PHUs) that cover the entire geography of the province. Although PHUS were commissioned with leading the distribution of the COVID-19 vaccine within their respective areas [@rotenberg2021], the vaccine rollout occurred with significant interaction between PHO and OH; in many instances both agencies had to work together to organize vaccination clinics, with personnel associated with OH also being actively involved when demand exceeded the capacity of PHO personnel, especially in rural areas [@ashcroft2023]. Indeed, based on the experience of COVID-19, there is an ongoing debate on how OH and PHO will interact in the future, and the challenges that this will entail for the healthcare system of the province [@dion2022].

Therefore, considering the relatively recent adoption of the Health Region model and its alignment with the onset of the COVID-19 pandemic, there is an ongoing need to analyze the existence of geographical disparities in vaccination uptake within the Health Regions and identify the socio-demographic groups that might be affected, as this can serve as an indirect assessment of the state of the implementation of the new model while helping identify ongoing challenges that decision-makers might need to address to ensure the long-term  success of this model and its interaction with PHO, which is specially important considering that previous research has highlighted disparities in the level of activity of each Health Region [@sethuram2023].

<!-- before the advent of a future pandemic -->

<!-- In the context of Canada's pandemic response goals, which have been to minimize serious illness and deaths while minimizing societal disruption [@naci2023], there is an ongoing need to analyze the existence of geographical disparities in vaccination uptake within Ontario and identify the socio-demographic groups that might be affected, as this can help identify inequalities that might exist and that need to be addressed before the advent of a future pandemic. -->

<!-- there is a need to analyze if there are ongoing disparities in health access under this approach that need to be addressed before they are exploited by a new disease or public health threat. In this regard, previous research has highlighted disparities in the level of activity of each Health Region [@sethuram2023]. Therefore, analyzing differences in vaccination uptake within the Health Regions and can help identify which socio-demographic groups are the most vulnerable and what areas of the province deserve special attention by decision-makers.  -->

<!-- Due to the relatively recent adoption of the Health Region model and its alignment with the onset of the COVID-19 pandemic,  -->

<!-- In the context of Canada's pandemic response goals, which have been to minimize serious illness and deaths while minimizing societal disruption [@naci2023], there is an ongoing need to analyze the existence of geographical disparities in vaccination uptake within Ontario and identify the socio-demographic groups that might be affected, as this can help identify inequalities that might exist and that need to be addressed before the advent of a future pandemic. -->

<!-- how the relatively recent adoption of the Health Region model the existence of geographical disparities in vaccination uptake within the Canadian provinces and to identify the socio-demographic groups affected, as this can help identify inequalities that might exist and that need to be addressed before the advent of a future pandemic. -->

<!-- there is a need to analyze if there are ongoing disparities in health access under this approach that need to be addressed before they are exploited by a new disease or public health threat. In this regard, previous research has highlighted disparities in the level of activity of each Health Region [@sethuram2023]. Therefore, analyzing differences in vaccination uptake within the Health Regions and can help identify which socio-demographic groups are the most vulnerable and what areas of the province deserve special attention by decision-makers.  -->

<!-- However, in the context of Canada's pandemic response goals, which have been to minimize serious illness and deaths while minimizing societal disruption [@naci2023], there is an ongoing need to analyze the existence of geographical disparities in vaccination uptake within the Canadian provinces and to identify the socio-demographic groups affected, as this can help identify inequalities that might exist and that need to be addressed before the advent of a future pandemic. -->

<!-- there is a need to analyze if there are ongoing disparities in health access under this approach that need to be addressed before they are exploited by a new disease or public health threat. In this regard, previous research has highlighted disparities in the level of activity of each Health Region [@sethuram2023]. Therefore, analyzing differences in vaccination uptake within the Health Regions and can help identify which socio-demographic groups are the most vulnerable and what areas of the province deserve special attention by decision-makers.  -->

Therefore, in this study we hypothesized that there were differences in COVID-19 vaccination rates between the Health Regions between October 2021 and January of 2022. To understand and contextualize these differences, we by included socio-economic factors in our analysis, aiming to identify in which demographic groups were particularly impacted, in order to provide an assessment of the current state of healthcare access in Ontario.

### 2023-07-02 manuscript_v6/02-methods.qmd
updated methods

We used data from the *Survey of COVID-19 related Behaviours and Attitudes*, a repeated cross sectional survey focused on the Canadian province of Ontario that was commissioned by the Fields Institute for Research in Mathematical Sciences and the  Mathematical Modelling of COVID-19 Task Force under ethical guidance from the University of Toronto (under protocol 00043317), and which ran between September 30th, 2021 and January 17th,2022. The survey collected socio-economic information from participants (@tbl-descriptive-stats), their location (nearest municipality,  as shown in @fig-map), the date of access to the survey, and asked information on vaccination status by using the question "Have you received the first dose of the COVID vaccine?", with possible answers "yes" and "no". The original dataset contained 39,029 observations.

By design, the survey allowed respondents to exit at any time and deployed the questions randomly. This resulted in $\approx$ 84\% of the observations having multiple missing answers or being incomplete due to participants leaving the survey at different stages. Therefore, we selected 6,343 observations that were labelled as "complete" in the dataset and that had answers for all covariates considered in our analysis. Later, we matched the city of each observation with its corresponding LHIN and Health Region, and removed observations from areas with low representation (254 observations corresponding to the North West and North East Health Regions). After all the preliminary analyses, the total number of observations used for analysis was 6,236 and included the East, Central, Toronto, and West Health Regions covering between October 1st,2021 and January 17, 2022. The original dataset, clean dataset, and details on the data cleaning process are described in detail in the [GitHub repository](https://github.com/aimundo/Fields_COVID-19/) for this paper.

>>>>

We used data from the *Survey of COVID-19 related Behaviours and Attitudes*, a repeated cross sectional survey focused on the Canadian province of Ontario that was commissioned by the Fields Institute for Research in Mathematical Sciences and the  Mathematical Modelling of COVID-19 Task Force under ethical guidance from the University of Toronto (under protocol 00043317), and which ran between September 30th, 2021 and January 17th,2022.

Briefly, the survey was deployed using random domain intercept technology, a methodology where internet users whose device meta-data indicated their presence in the province of Ontario had a random chance of being redirected to the survey after they had clicked on a registered but commercially inactive web link, or typed in a web address for a site that was dormant but that was temporarily managed by the company that administered the survey (RIWI Corp) [@sargent2022]. Users then decided whether to anonymously participate in the survey, and those that participated were able to exit the survey at any time. After the survey closed, regardless if it was complete or incomplete, access was denied to any further users with the same internet protocol address (IP), effectively allowing each user only one opportunity to participate in the survey. Users who indicated they were under the age of 16 were exited from the survey without creating a record. Finally, the personal identifier information from each user that participated in the survey was automatically scrubbed and replaced by a unique ID.

The survey allowed users to enter their socio-economic information, and asked information on vaccination status by using the question "Have you received the first dose of the COVID vaccine?", with possible answers "yes" and "no" (@tbl-descriptive-stats). Additionally, the survey automatically collected the geographical location of the respondent (using the nearest municipality,  as shown in @fig-map), and the date of access to the survey. The original dataset contained 39,029 observations with a response rate of 16.25% (observations where users had complete answers), due to the survey design that allowed respondents to exit at any time and deployed the questions randomly. We selected these  observations with complete answers (6,343 observations) for our analysis. Later, we matched the city of each observation with its corresponding LHIN and Health Region, and removed observations from areas with low representation (254 observations corresponding to the North West and North East Health Regions). After all the preliminary analyses, the total number of observations used for analysis was 6,236 and included the East, Central, Toronto, and West Health Regions covering between October 1st,2021 and January 17, 2022. The original dataset, clean dataset, and details on the data cleaning process are described in detail in the [GitHub repository](https://github.com/aimundo/Fields_COVID-19/) for this paper.
